Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-13 aistats.tex

\section{Optimization Perspective}
\label{sec:optimization}
As we saw in the technical background, MAP can be interpreted as stochastic mirror descent (SMD).
This means that \textbf{1)} we may obtain a convergence rate for MAP from an optimization analysis, and \textbf{2)} any insights gained from the MAP may inform further design and analysis of SMD.
We first review the assumptions of relative smoothness, useful to deal with non-smooth functions, before investigating 3 analysis of SMD with the MAP.
\rlp{mention that examples have so far been lacking for relative SMD.}

\subsection{Relative Smoothness}
Mirror descent (MD)\footnote{
Also known as
Bregman (proximal) gradient, relative gradient descent or NoLips.
}
\citep{nemirovski1983problem,beck2003mirror}
%Nemirovski introduced the algorithm, and Beck established the connection with Bregman divergences, e.g. the argmin form.
and SMD 
\citep{nemirovski2009robust,ghadimi2012optimal} 
are typically encountered in non-smooth (online) optimization,
under bounded (or Lipschitz) gradient assumption on the objective $f$
and strong convexity assumption on the potential $\logpart$
\citep[Th.4.2(MD) \& Th.6.3(SMD)]{bubeck2015convex}.
In our case, none of these assumptions may hold. 
For instance $\logpart = -\log$ is neither smooth nor strongly convex.

Recently, these assumptions have been relaxed to allow for non-smooth functions $f$
\footnote{$f$ is $L$-smooth if its gradient is $L$-Lipschitz}.
and non-strongly convex potential $\logpart$, first in the deterministic setting
\citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively}, then in the stochastic setting \citep{hanzely2018fastest, dragomir2021fast, dorazio2021stochastic}.
These assumptions have been replaced with
the $\mu$-strong convexity and $L$-smoothness of $f$
\emph{relative} to $\logpart$, defined as
\aligns{
	\mu \cB_{A}(x, y)
	\leq
	\cB_f(x,y)
	\leq
	L \cB_A(x,y) \; .
}
When $\logpart = \norm{\cdot}^2$, we recover standard (Euclidean) smoothness, and gradient descent. 
Those conditions ensure the linear convergence rate of MD with $A$ as the reference function.
Exponential families log-likelihood perfectly fit into this framework, as
\aligns{
	f(\theta) = A(\theta) - \expect{\lin{X, \theta}}
}
is $1$-smooth and $1$-strongly convex relative to $A$.

\subsection{Bounding the Randomness}
Smoothness is not the only issue.
To analyze stochastic algorithms, one also need to quantify the randomness.
Many assumptions exist for SGD \citep[\S3 for a modern review]{khaled2020better}.
We now review attempts at analyzing SMD under relative smoothness assumption, in the light of  the MAP example.
See \cref{tbl:assumptions} for a summary.

>>>>

\section{Optimization Perspective}
\label{sec:optimization}

%This section show how MAP can be interpreted as stochastic mirror descent (SMD). Thanks to this strong link, we show that \textbf{1)} we may obtain a convergence rate for MAP from the rate od SMD, and \textbf{2)} any insights gained from the MAP may inform on further design and analysis of SMD.
%Then, we review the assumptions of relative smoothness, useful to deal with non-smooth functions, before investigating three different analyses of SMD with the MAP.
%\rlp{mention that examples have so far been lacking for relative SMD.}
%
%\subsection{MAP as Stochastic Mirror Descent}
%\label{ssec:MAP=SMD}
%By the Bayes rule, the posterior can always be written iteratively $p(\nat | \mD_n) \propto p(X_n | \nat) p(\nat | \mD_{n-1})$.
%Plugging in~\eqref{eq:joint_likelihood}, and introducing $f_n(\nat) = - \log p(X_{n} | \nat)$, we see that the MAP verifies
%\begin{align*}
%	\hat \nat_{n}
%	= \argmin_\nat f_n(\nat) + (n_0 + n -1) \bregman(\nat; \hat \nat_{n-1})
%    %\label{eq:SBPP}
%\end{align*}
%which is the update of a Stochastic Bregman Proximal Point update with step-size $\inv{n_0 + n - 1}$.
%%Maybe skip this SBPP story altogether
%If we expand $f_n(\nat) = \logpart(\nat) - T_n$,  we get (after some manipulation with the Bregman divergence)
%\begin{align*}
%		\hat \nat_{n}
%	= \argmin_\nat \langle g_{n}(\hat \nat_{n-1}), \nat \rangle + (n_0 + n) \bregman(\nat; \hat \nat_{n-1})
%    %\label{eq:SMD}
%\end{align*}
%where $g_{n}(\nat) := \nabla f_n(\nat) = \hat \m_{n-1} - T_n$.
%We recognize the update of Stochastic Mirror Descent (a.k.a. Stochastic Bregman Gradient) with step-size $\lr_n = \inv{n_0 + n}$.
%From the first order optimality condition, we see this is a stochastic gradient step in the dual $\hat \m_n = \hat \m_{n-1} - \lr_n g_n$.
%This parallel means we could obtain a finite convergence rate for the  MAP from an analysis of these algorithms.
%In particular, it could potentially apply to a wide range of distribution $\cD$ on $X$.
%Note that this optimization perspective does not hold for the MLE %($n_0=0$)
%as its implicit uniform prior may correspond to an initialization outside of $\Theta$.

As we saw in the technical background, MAP can be interpreted as stochastic mirror descent (SMD).
This means that \textbf{1)} we may obtain a convergence rate for MAP from an optimization analysis, and \textbf{2)} any insights gained from the MAP may inform further design and analysis of SMD.
We first review the assumptions of relative smoothness, useful to deal with non-smooth functions, before investigating 3 analysis of SMD with the MAP.
\rlp{mention that examples have so far been lacking for relative SMD.}

\subsection{Relative Smoothness}
Mirror descent (MD) \citep{nemirovski1983problem,beck2003mirror}, also known as
Bregman (proximal) gradient, relative gradient descent or NoLips,
%Nemirovski introduced the algorithm, and Beck established the connection with Bregman divergences, e.g. the argmin form.
and SMD 
\citep{nemirovski2009robust,ghadimi2012optimal} 
are typically encountered in non-smooth (online) optimization,
under bounded (or Lipschitz) gradient assumption on the objective $f$
and strong convexity assumption on the potential $\logpart$
\citep[Th.4.2(MD) \& Th.6.3(SMD)]{bubeck2015convex}.
In our case, none of these assumptions may hold. 
For instance $\logpart = -\log$ is neither smooth nor strongly convex.

Recently, these assumptions have been relaxed to the $\mu$-strong convexity and $L$-smoothness of $f$
\emph{relative} to $\logpart$, defined as
\aligns{
	\mu \cB_{A}(x, y)
	\leq
	\cB_f(x,y)
	\leq
	L \cB_A(x,y) \; .
}
This extends the theory to a family of non-smooth functions $f$ and non-strongly convex potential $\logpart$, first in the deterministic setting
\citep{birnbaum2011distributed, bauschke2017descent, lu2018relatively}, then in the stochastic setting \citep{hanzely2018fastest, dragomir2021fast, dorazio2021stochastic}. When $\logpart = \norm{\cdot}^2$, we recover standard (Euclidean) smoothness, and gradient descent. 
Those conditions ensure the linear convergence rate of MD with $A$ as the reference function.

\textcolor{red}{Sure of this formulation?} Exponential families log-likelihood perfectly fit into this framework, as
\aligns{
	f(\theta) = A(\theta) - \expect{\lin{X, \theta}}
}
is $1$-smooth and $1$-strongly convex relative to $A$. However, the (relative) smoothness is not biggest issue in our convergence analysis, but the bound on the variance, as explained in the next section.

\subsection{Bounding the Randomness}
To analyze stochastic algorithms, one also need to quantify the randomness.
Many assumptions exist for SGD \citep[\S3 for a modern review]{khaled2020better}.
We now review attempts at analyzing SMD under relative smoothness assumption, in the light of  the MAP example.
See \cref{tbl:assumptions} for a summary.
